---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r echo=TRUE}
data <- read.csv("activity.csv")
str(data)
```

## What is mean total number of steps taken per day?
```{r echo=TRUE}
byday <- tapply(data$steps, data$date, sum)
hist(byday, main = "Histogram of Total Steps per Day, including NAs",
     xlab = "Steps")
summary(byday)
```


## What is the average daily activity pattern?
```{r echo=TRUE}
interval <- tapply(data$steps, data$interval, mean, na.rm = TRUE)
timeaverage <- data.frame(interval = names(interval), steps = interval,
                          row.names = NULL)
with(timeaverage, plot(interval, steps, type = "l", 
                                   main = "Time Series, NAs excluded",
                                   ylab = "average steps",
                                   xlab = "5-minute interval"))
timeaverage[which.max(timeaverage$steps),]  
```
The time interval at 835 minutes has the highest step value averaged across all 
days (~206 steps).

## Imputing missing values
```{r echo=TRUE}
nasteps <- is.na(data$steps)
sum(nasteps)
```
There are 2304 missing 'step' values in the data set. These can be imputed using
the overall mean step value (37.3826).

```{r echo = TRUE}
data2 <- data
data2$steps[is.na(data2$steps)] <- mean(data2$steps, na.rm = TRUE)

byday2 <- tapply(data2$steps, data2$date, sum)

hist(byday2, 
     main = "Histogram of Total Steps per Day, NAs = mean steps",
     xlab = "Steps")
summary(byday2)
```
Imputing the missing values with the mean step count has a negligible impact on 
the data as a whole. The median value changes slightly. 

## Are there differences in activity patterns between weekdays and weekends?
```{r echo = TRUE}
data2$date <- as.Date(data2$date)
data2$day <- weekdays(data2$date)

weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")

data2$wkend <- ifelse(data2$day %in% weekdays, "weekday", "weekend")
weekdaysteps <- data2[data2$wkend == "weekday",]
weekendsteps <- data2[data2$wkend == "weekend",]

wkday_interval <- tapply(weekdaysteps$steps, weekdaysteps$interval, mean, 
                         na.rm = TRUE)

wkend_interval <- tapply(weekendsteps$steps, weekendsteps$interval, mean, 
                         na.rm = TRUE)

wkday_intervaldata <- data.frame(interval= names(wkday_interval),
                                 steps= wkday_interval,
                                 day= "weekday", row.names = NULL)
wkend_intervaldata <- data.frame(interval= names(wkend_interval),
                                 steps= wkend_interval,
                                 day= "weekend", row.names = NULL)

data3 <- rbind(wkday_intervaldata, wkend_intervaldata)
data3$interval <- as.numeric(data3$interval)
data3 <- data3[order(data3$day),]
library(lattice)
xyplot(steps ~ interval | day, data=data3, type = "l", layout = c(1,2))
```

